White Wine Exploration by Divya Ramanujachari
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages used in the analysis
library(ggplot2)
library(gridExtra)
library(reshape2)
library(GGally)
library(tidyr)
library(dplyr)
library(lattice)
library(MASS)
library(memisc)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
wineData <- read.csv("wineQualityWhites.csv")
```

This report analyses the results of physicochemical tests on 4898 samples of 
white wine with a view toward determining the factors that impact the wine 
quality.

# Univariate Plots Section

### Dimensions of the dataset

```{r echo=FALSE, message=FALSE, warning=FALSE, Dimensions}
dim(wineData)
```

### Fields in the dataset

```{r echo=FALSE, message=FALSE, warning=FALSE, Headers}
str(wineData)
```

### Preprocessing

```{r echo=FALSE, message=FALSE, warning=FALSE, Cleanup}
# Converting the quality attribute to a factor variable
wineData$quality <- as.factor(wineData$quality)

# Dropping the serial number attribute
wineData <- subset(wineData, select=-1)
```

Based on the attribute description, I grouped the attributes into four sets
for ease of analysis. These are:

- Factors related to acidity (fixed.acidity, volatile.acidity, citric.acid)

- Factors related to sulfur dioxide (free.sulfur.dioxide, volatile.sulfur.dioxide,
sulphates)

- Factors related to salinity (pH, chlorides)

- Factors related to density (density, residual.sugar, alcohol)

The last category is directly related to the statement from the description that
says "density depends on the percent alcohol and sugar content."

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot_Function}
univariate_analysis <- function(variable, name, bin_size, upper_limit){
  summary_stats <- summary(variable)
  summary_data <- paste(names(summary_stats), format(summary_stats, digits = 2),
                        collapse = "; ")
  ggplot(aes(x = variable), data = wineData) +
    geom_histogram(aes(y = ..density..), binwidth = bin_size) +
    geom_density(colour=c("blue")) +
    scale_x_continuous(limits=c(min(variable), upper_limit)) +
    labs(x=paste(name), y="Density") +
    ggtitle(paste(name), subtitle=summary_data)
}
```

###Factors related to acidity

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot_1}
univariate_analysis(wineData$fixed.acidity, "Fixed acidity", 0.1, 10.5)
univariate_analysis(wineData$volatile.acidity, "Volatile acidity", 0.01, 0.75)
univariate_analysis(wineData$citric.acid, "Citric acid", 0.01, 0.8)
```

The fixed acidity appears to be distributed normally around the median of 6.8. 
Outliers are present, as shown by the gap between the third quantile and 
maximum value. The volatile acidity is positively skewed with a long tail. The 
reason for this might be that, according to the description, high levels of 
volatile acids affect the wine taste. The citric acid values are mostly normal, 
with an intriguing peak near 0.5. 

###Factors related to sulfur dioxide

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot_2}
univariate_analysis(wineData$free.sulfur.dioxide, "Free sulfur dioxide", 1, 100)
univariate_analysis(wineData$total.sulfur.dioxide, "Total sulfur dioxide", 
                          1, 275)
univariate_analysis(wineData$sulphates, "Sulphate", 0.01, 0.9)
```

The free sulfur dioxide distribution appears to be long-tailed; please note that
outliers, not shown in the plot, exist sporadically till a max of 289. According 
to the variable description document this is used as a preservative. The total
sulfur dioxide also shows some high values; this is said to affect the quality 
of wines and must be investigated further. Described as an antimicrobial and
antioxidant, sulphates are somewhat normally distributed but values in the
higher range are also observed.

###Factors related to salinity

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot_3}
univariate_analysis(wineData$pH, "pH", 0.01, 3.8)
univariate_analysis(wineData$chlorides, "Chlorides", 0.005, 0.1)
```

The pH values show a normal distribution and conform to the prescribed range
of 3-4. The chlorides too exhibit a normal distribution with a few wines having
a greater amount of salt.

###Factors related to density

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plot_4}
univariate_analysis(wineData$density, "Density", 0.0005, 1.04)
univariate_analysis(wineData$residual.sugar, "Residual sugar", 0.1, 20)
univariate_analysis(wineData$alcohol, "Alcohol", 0.1, 14)
```

The density values do not show much variability, a couple of outliers are 
observed very far from the mean. The residual sugar is positively skewed and
has a very long tail. The values cover a large range from 0.6 to 65.8. This is
also the trend of the alcohol value distribution. There is a spike at 9.5, which
is below the median and mean values.

###Quality attribute

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Summary}
univariate_analysis(as.numeric(wineData$quality), "Quality", NA, NA)
```

This graph indicates how the dataset is not well balanced in terms of wine
quality; there are many wines in the mediocre range but very few that can be
labelled as very bad or very good.

# Univariate Analysis

### What is the structure of your dataset?

The dataset comprises 4898 observations with 13 variables, of which 11 are 
physicochemical properties and one (quality) quantifies a sensory parameter. The
remaining one is an identifier. A detailed description of the attributes was 
provided with the dataset. 

### What is/are the main feature(s) of interest in your dataset?

As mentioned, the objective of the study is to determine the factors that 
influence quality and this is the main feature of interest. However, the provided
dataset is not well balanced as most of the wines are of mediocre
quality (rating of 5 to 7) and there are only a few examples of very bad or
excellent wines. This might provide an incomplete picture.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

While answering this would require bivariate analysis with regard to quality,
from the description, factors like volatile acidity, citric acid, total sulfur 
dioxide, and residual sugar have a direct effect on the taste. I will be 
investigating these in detail next.

### Did you create any new variables from existing variables in the dataset?

I did not create any new variable.

### Of the features you investigated, were there any unusual distributions?

In the citric acid value distribution there is an intriguing peak near 0.5, for
which I could not find any reasonable explanation. Similarly there is a peak in
the alcohol distribution at a value much below the mean/median, which is not
in focus because of its very long tail.

### Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

I dropped the variable X, which was only an identifier variable and not relevant
to the investigation. I also cast the quality into a factor variable. Initially 
it was an integer variable, which was inconsistent with its description as a 
categorical variable that can take only specific values.

# Bivariate Plots Section

In this section, I first analysed the correlation between the various variables
and the quality factor. Following this, the correlations between the variables
were analyzed, for which I considered an absolute value of more than 0.5 as being
noteworthy.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot_Function}
bivariate_analysis <- function(variable, name){
  p1 <- ggplot(aes(y = variable, x = quality), data = wineData) +
    geom_point(alpha = 0.5) +
    stat_summary(fun.y="mean", geom="line", color="red", aes(group=1)) +
    ggtitle("Scatter plot")
  p2 <- ggplot(data=wineData, aes(y=variable, x=factor(0), fill=quality)) +
    geom_boxplot() +
    ggtitle("Boxplot")
  grid.arrange(p1,p2, top=paste(name, "vs. quality: corr", 
                                round(cor(variable,as.numeric(wineData$quality))
                                      ,2)),ncol=2)
}
```

###Factors related to acidity

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot_1}
bivariate_analysis(wineData$fixed.acidity, "Fixed acidity")
bivariate_analysis(wineData$volatile.acidity, "Volatile acidity")
bivariate_analysis(wineData$citric.acid, "Citric acid")
```

The correlation of fixed acidity, volatile acidity, and citric acid with quality
was computed to be -0.11, -0.19, and -0.01. Looking at the scatterplot and 
boxplots I was not able to discern any strong pattern across the quality values.

###Factors related to sulfur dioxide

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot_2}
bivariate_analysis(wineData$free.sulfur.dioxide, "Free sulfur dioxide")
bivariate_analysis(wineData$total.sulfur.dioxide, "Total sulfur dioxide")
bivariate_analysis(wineData$sulphates, "Sulphate")
```

The correlation of free sulfur dioxide, total sulfur dioxide, and sulphates with
quality was computed to be 0.01, -0.17, and 0.05. As in the previous set, I was 
not able to come to any conclusion about variation in quality from the plots. 

###Factors related to salinity

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot_3}
bivariate_analysis(wineData$pH, "pH")
bivariate_analysis(wineData$chlorides, "Chlorides")
```

The correlation of pH and chlorides with quality was computed to be 0.1 and
-0.21. From the plots, it was observed that higher quality wines tend to have
higher pH on an average. Similarly higher quality wines tended to have lower 
chlorides on an average. However, these observations exclude the outliers and
there are not enough samples in the bad and excellent categories for this
observation to be significant.

###Factors related to density

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot_4}
bivariate_analysis(wineData$density, "Density")
bivariate_analysis(wineData$residual.sugar, "Residual sugar")
bivariate_analysis(wineData$alcohol, "Alcohol")
```

The correlation of density, residual sugar, and alcohol with quality was 
computed to be -0.31, -0.1, and 0.44. No clear trend was observed in the plots
for residual sugars. An inverse trend was observed in the relation between the
density and quality. Further, the alcohol level and quality appear to be 
positively correlated; in general, wines of better quality have higher levels of 
alcohol.

###Correlations between variables other than quality

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot_5}
ggcorr(subset(wineData, select = -c(quality)), label = 'True')
```

The matrix shows that the correlation between alcohol and density and residual 
sugar and density was found to be highest at 0.8. This was followed by the
correlation between free sulfur dioxide and total sulfur dioxide, which was 0.6.
Next, scatter plots are plotted to examine the linear relationship between the 
various variables and fit and approximate regression line.

###Plot of density vs. alcohol

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot_6}
ggplot(aes(y = density, x = alcohol), data = wineData) +
  geom_point(alpha=0.5) +
  scale_y_continuous(limits=c(NA,1.005)) +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle(paste("Density vs. alcohol; corr:",
                round(cor(wineData$density,wineData$alcohol),2)))
```

The correlation value is reflected by the good fit of the regression line; the
relationship between density and alcohol is worth exploring in detail.

###Plot of density vs. residual sugar

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot_7}
ggplot(aes(y = density, x = residual.sugar), data = wineData) +
  geom_point(alpha=0.2) +
  scale_x_continuous(limits=c(NA,25)) +
  scale_y_continuous(limits=c(NA,1.005)) +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle(paste("Density vs. residual sugar; corr:",
                round(cor(wineData$density,wineData$residual.sugar),2)))
```

The correlation value between density and residual sugar is highest at 0.84. 
This is evident from the scatter plot and it will be interesting to see how
quality varies with these parameters.

###Plot of free sulfur dioxide vs. total sulfur dioxide

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plot_8}
ggplot(aes(y = total.sulfur.dioxide, x = free.sulfur.dioxide), data = wineData) +
  geom_point(alpha=0.2) +
  scale_x_continuous(limits=c(NA,100)) +
  scale_y_continuous(limits=c(NA,300)) +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle(paste("Free sulfur dioxide vs. total sulfur dioxide; corr:",
                round(cor(wineData$free.sulfur.dioxide,
                          wineData$total.sulfur.dioxide),2)))
```

This pair was analyzed because of the high correlation value. However, 
considering that the free sulfur dioxide is a subset of the total sulfur dioxide 
according to the dataset, this comes as no surprise and this relationship will
not be assessed with respect to quality.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

In summary, quality was found to be most closely correlated with alcohol and 
density. However, even these relations were not very strong and inconsistencies 
in some wine quality categories with less number of representative wines 
were observed.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

The correlation between alcohol and density and residual sugar and density was 
found to be highest at 0.8. This was followed by a correlation of 0.6 between 
free sulfur dioxide and total sulfur dioxide. The last is evident because of
the subset-superset nature of the variables.

### What was the strongest relationship you found?
The strongest relationship was that between density and residual sugar. These
factors are positively correlated with a factor of 0.84.

# Multivariate Plots Section

From the previous section it was found that the relationship among quality,
density, alcohol, and residual sugar are good candidates for further 
investigation.

###Plot of density vs. alcohol by quality

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plot_1}
ggplot(aes(y = density, x = alcohol, colour=quality), data = wineData) +
  geom_point(alpha=0.5) +
  scale_y_continuous(limits=c(NA,1.0025)) +
  facet_wrap(~quality) +
  ggtitle(paste("Density vs. alcohol by quality"))
```

As seen previously, there is a negative correlation between the density and 
alcohol levels. When segregated by quality, it is observed that while low and 
mediocre quality wines have values across the entire range, high quality wines
typically have high alcohol content and minimum values of density.

###Plot of density vs. residual sugar by quality

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plot_2}
ggplot(aes(y = density, x = residual.sugar, colour=quality), data = wineData) +
  geom_point(alpha=0.5) +
  scale_x_continuous(limits=c(NA,25)) +
  scale_y_continuous(limits=c(NA,1.005)) +
  facet_wrap(~quality) + 
  ggtitle(paste("Density vs. residual sugar grouped by quality")) 
```

The positive correlation between the density and residual sugar is maintained 
across wines of all qualities. Furthermore, it is observed that high quality 
wines tend to have low values of density as well as residual sugar.

###Plot of alcohol vs. residual sugar by quality

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plot_3}
ggplot(aes(y = alcohol, x = residual.sugar, colour=quality), data = wineData) +
  geom_point(alpha=0.5) +
  scale_x_continuous(limits=c(NA,20)) +
  facet_wrap(~quality) + 
  ggtitle(paste("Alcohol vs. residual sugar grouped by quality")) 
```

This plot confirms the findings of the previous plots, namely, wines of high 
quality have high alcohol content and low residual sugar.

###Linear regression model

Density, residual sugar, and alcohol parameters were used to construct a linear
regression model for determining the quality of wines.

```{r echo=FALSE, message=FALSE, warning=FALSE, Regression_Model}
lmfit <- lm(I(as.numeric(quality)) ~ I(alcohol) + I(1/density) + 
              I(1/residual.sugar), data=wineData)
summary(lmfit)
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Plotting the various combinations of density, residual sugar, and alcohol by 
quality brought out features of high quality wines, specifically, 
high alcohol content, low residual sugar, and low density.

### Were there any interesting or surprising interactions between features?

The findings in this section reinforced the correlation between alcohol and 
density and residual sugar and density across all wine qualities. 

### Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

Yes, I created a linear regression model using the abovementioned variables. 
Unfortunately, these parameters proved to be inadequate to predict the variance 
in quality, the R-squared was only about 20%. It could be that the imbalance in
data induced errors in analysis.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
summary_stats <- summary(as.numeric(wineData$quality))
summary_data <- paste(names(summary_stats), format(summary_stats, digits = 2),
                      collapse = "; ")
ggplot(aes(x = as.numeric(quality)), data = wineData) + 
  geom_histogram(binwidth = 1) +
  scale_x_continuous(limits = c(0, 10), breaks=c(seq(0,10,1))) +
  labs(x="Quality", y="Count") +
  ggtitle("Distribution of wine quality", subtitle=summary_data)
```

### Description One

While this appears to be a very simple plot, it clearly reflects the shortcoming
of the study; specifically, the imbalance in the dataset that has great impact
on the reliability of the findings. As seen, wines are graded in quality
from 0 to 10. However, the data in the provided dataset has no information at 
all about the excellent wines and very little data on the terribly bad ones.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
ggplot(aes(y = density, x = alcohol, colour=quality), data = wineData) +
  geom_point(alpha=0.5) +
  scale_y_continuous(limits=c(NA,1.0025), name="Density (g/cm^3)") +
  scale_x_continuous(name="Alcohol (% by volume)") +
  facet_wrap(~quality) +
  ggtitle(paste("Density vs. alcohol by quality"))
```

### Description Two

I chose this plot as it clearly demonstrates one of the key findings of this 
study. Specifically, there is a negative correlation between the density and 
alcohol levels. Also, it is observed that, high quality wines typically have 
high alcohol content and minimum values of density.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(aes(y = density, x = residual.sugar, colour=quality), data = wineData) +
  geom_point(alpha=0.5) +
  scale_x_continuous(limits=c(NA,25), name="Residual sugar (g / dm^3)") +
  scale_y_continuous(limits=c(NA,1.005), name="Density (g/cm^3)") +
  facet_wrap(~quality) + 
  ggtitle(paste("Density vs. residual sugar grouped by quality"))
```

### Description Three

This plot shows that a positive correlation exists between the density and 
residual sugar across wines of all qualities. Furthermore, it is observed that 
high quality wines tend to have low values of density as well as residual sugar.

# Reflection

In this project, I performed an exploration of physicochemical parameters of 
4898 samples of white wine to analyze whether any of the parameters have a direct
effect of the quality. It was found that higher quality wines did exhibit some
characteristic features of low density, low residual sugar, and high alcohol
content. However, some lower quality wines too possess the same characteristics.
Further analysis is needed to establish a more comprehensive model. This could
be achieved by adding more data points and taking into consideration other
features as well.